Updates a Procrustes and a PCA

### 2025-05-02 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@

## Running the pca on the tangent space data
original_pca <- prcompfast(original_data$tan*-1) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST

@@

>>>>

# The pseudo math:

In pseudo math, the whole procedure is something like:

 1. Procrustes = multiplying the landmark data `D` by some estimated transformation matrix `M1`.
 2. PCA = multiplying the procrustes data `M1*D` by some transformation matrix (the eigvenctors) `M2`.

So we should technically be able to get the numerical values from both transformation matrices and use them to feed in the new data to them to have something like `M2*M1*D'` where `D'` are the new landmarks.
In that case if `M2*M1*D = traitspace`, then if `D' = D`, we should get the exact same results (a traispace with _n_ species with all the same coordinates).
And if `D' = D + x` where `x` are _m_ new specimens, we should have a traitspace with the exact same results for `D` and new results of `x` that are "correctly" positioned into that space.

@@

## Running the pca on the tangent space data
original_pca <- prcompfast(original_data$tan*-1)

@@

```{r, eval = FALSE, echo = FALSE}
orp <- function(A, mshape=NULL) { 
  p <- dim(A)[1]
  k <- dim(A)[2]
  n <- dim(A)[3]
  if (is.null(mshape))
      mshape <- arrMean3(A)

  m.size <- cSize(mshape)
  Xc <- as.vector(mshape/m.size)
  X <- vecx(A)/m.size
  ##direction along mshape onto plane
  XtoPlane <- t(apply(X,1,function(x){x <- t(c(crossprod(x,Xc))*Xc)}))
  X1 <- X-XtoPlane
  X1 <- t(X1)+Xc
  proj <- array(X1, dim=c(p, k, n))*m.size
  return(proj)
}

## Testing align2procSym without changing the mean
new_individuals <- testing_projection_shapes
original_data <- whole_gpa
new_landmarks <- new_individuals

## Internal logic for align2procSym
n <- dim(new_landmarks)[3]
atts <- attributes(whole_gpa)
newdatarot <- new_landmarks

## Centroid Sizing
mysize <- apply(new_landmarks,3,cSize)
for (i in 1:n){
    new_landmarks[,,i] <- new_landmarks[,,i]/mysize[i]
}

for (i in 1:n) {
    newdatarot[,,i] <- rotonto(x = whole_gpa$mshape, y =new_landmarks[,,i],scale=atts$scale,reflection=atts$reflect,centerweight=atts$centerweight,weights=atts$weights)$yrot
}

orpdata <- orp(newdatarot,whole_gpa$mshape)
noorpda <- newdatarot

## New data needs to be projected on the tangent space because of symmetry.
symtan_orpda <- sweep(orpdata, 1:2, whole_gpa$mshape)
symtan_noorp <- sweep(noorpda, 1:2, whole_gpa$mshape)
proctan_ordat <- vecx(symtan_orpda)
proctan_norop <- vecx(symtan_noorp)

## Running the pca on the tangent space data
original_pca <- prcompfast(original_data$tan*-1)

## Getting the new PCAs
new_pca_orpda <- stats::predict(original_pca, newdata = proctan_ordat)#array.to(new_proc_tan, to = "matrix"))
new_pca_norop <- stats::predict(original_pca, newdata = proctan_norop)#array.to(new_proc_tan, to = "matrix"))

## Combine the new and old matrices for the output PCA
updated_pca <- original_pca$x*-1

plot(whole_gpa$PCscores[,c(1,2)], col = "grey", main = "Sanity check")
points(updated_pca[, c(1,2)], pch = 19, cex = 0.5, col = "grey")
points(whole_gpa$PCscores[dimnames(testing_projection_shapes)[[3]], c(1,2)], pch = 21, col = "darkgreen")
points(new_pca_orpda[, c(1,2)], col = "blue", pch = 19, cex = 0.5)
points(new_pca_norop[, c(1,2)], col = "orange", pch = 19, cex = 0.5)

```
